---
output:
  md_document:
    variant: markdown_github
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# dbloadss

This package and accompanying scripts is to test load times to MS SQL Server 17 (SS) from various methods:

1. Push from R with `RODBC`
2. Push from R with `odbc`
3. Pull from SS with a stored procedure

This package is really for number 3. We can do the rest from a script, although it's a useful place to keep some functions.

# Test Database

The code for this post runs on my Windows 10 laptop, where I have a local SQL Server 17 instance running, with a database called `ml`.

# Load Times

We're going to push a big data frame into SQL Server using three methods. The data set is a simple randomly generated data frame.

```{r rand_data}
library(dbloadss)

random_data_set(n_rows = 5, n_cols = 5)
```

## RODBC

`RODBC` was, for a long time, the standard way to connect to SQL Server. 

## ODBC

`odbc` is a relatively new package from RStudio which provides a DBI compliant ODBC interface.

## SQL Server External Script

An alternative approach is to use the new features in SQL Server 17 (and 16) for calling out to R scripts from SQL. This is done via the `sp_execute_external_script` command, which we will wrap in a stored procedure.

# License 

MIT © Mango Solutions
